---
title: "Remote sources: storage, authentication and windows"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

```{r setup}
library(cptkirk)
```

cptkirk reads Cloud-Optimised GeoTIFFs straight from cloud object stores --- S3,
Google Cloud Storage, Azure, plain HTTP(S), or local disk. This article covers
the three things people ask first: how a source is addressed, how it
authenticates, and how to pull pixels (a window, or a whole reprojected AOI)
into R.

We use three public datasets, chosen to span the awkward cases:

| Dataset | Access | CRS | Type |
|---|---|---|---|
| Alpha Earth Foundations embeddings | plain `https://` (no auth) | UTM 36S | Int8, 64 bands |
| Sentinel-2 L2A (AWS Open Data) | `s3://` (anonymous + region) | UTM | Byte / UInt16, one COG per band |
| GEBCO 2024 bathymetry | plain `https://` (no auth) | EPSG:4326 | Int16, global |

## A source is a single URL

Where libraries like `obstore` / `async-tiff` separate a *store* (bucket,
region, credentials) from a *path* (the object key), cptkirk takes one
self-contained URL and works out the rest. The scheme (and, for AWS, the host)
selects the backend; the rest is the object key.

```{r aef}
aef <- paste0(
  "https://data.source.coop/tge-labs/aef/v1/annual/2021/36S/",
  "xekh5rjs4wg6wb9b4-0000000000-0000000000.tiff"
)
cog_info(aef)
```

`cog_info()` reads only the header and IFDs --- no pixels --- so it is a cheap
way to understand a file before fetching anything.

## Authentication comes from the environment

cptkirk never takes credentials as arguments; it reads them from the
environment, so secrets stay out of your scripts. It honours the **same
variables as GDAL's `/vsi*` drivers**, so a setup that already works with GDAL
needs no change: `AWS_ACCESS_KEY_ID` / `AWS_SECRET_ACCESS_KEY` / `AWS_REGION`,
`AWS_NO_SIGN_REQUEST=YES` for public buckets, `AWS_S3_ENDPOINT` for a custom
endpoint, `GOOGLE_APPLICATION_CREDENTIALS`, `AZURE_STORAGE_ACCOUNT`, and so on.
The object_store-native spellings (`AWS_SKIP_SIGNATURE`, ...) work too; where the
two differ, cptkirk translates the GDAL name. Only non-secret knobs are
translated R-side -- credentials pass straight from the environment to the
reader.

Mapping from the `obstore` store object:

| `obstore` | cptkirk |
|---|---|
| `S3Store("bucket", region = "us-west-2", skip_signature = True)` | `Sys.setenv(AWS_REGION = "us-west-2", AWS_SKIP_SIGNATURE = "true")` (or GDAL's `AWS_NO_SIGN_REQUEST = "YES"`) |
| `path = "key/within/bucket"` | fold into the URL: `s3://bucket/key/within/bucket` |
| `GeoTIFF.open(path, store = store)` | `cog_info(url)` / `ck_read(url)` / `ck_warp(url, dst)` |

Two situations:

* **Plain `https://` on a non-AWS host** (the AEF tile above, GEBCO below) uses
  the HTTP backend and needs nothing.
* **AWS-hosted data** --- whether you write `s3://…` *or*
  `https://….amazonaws.com/…` --- routes to the S3 backend, which signs requests.
  For a public bucket, switch signing off and set the region:

```{r sentinel-auth}
Sys.setenv(AWS_REGION = "us-west-2", AWS_SKIP_SIGNATURE = "true")

scene <- paste0(
  "s3://sentinel-cogs/sentinel-s2-l2a-cogs/12/S/UF/2022/6/",
  "S2B_12SUF_20220609_0_L2A/"
)
cog_info(paste0(scene, "TCI.tif"))
```

The selection is by **host**, not scheme: a `*.amazonaws.com` URL is treated as
S3 even over `https`, so it still needs the no-sign flag here. Either spelling
works --- `AWS_SKIP_SIGNATURE = "true"` (object_store) or
`AWS_NO_SIGN_REQUEST = "YES"` (GDAL); cptkirk normalises the GDAL `YES`/`NO` form
for you.

## Reading a window into R

`ck_read()` is `ck_warp()` that returns a base-R array instead of writing a
file. With no `t_srs` or `tr` it reads the native window directly (no GDAL); the
returned matrix/array carries `geotransform`, `crs` and `nodata` attributes, so
it is self-describing.

If you think in source **pixel offsets** (like async-tiff's
`Window(col_off, row_off, width, height)`), convert them to a target extent with
the geotransform and hand that to `ck_read()`:

```{r window}
gt <- cog_info(paste0(scene, "B04.tif"))$geotransform

col_off <- 4096L; row_off <- 2048L; width <- 512L; height <- 512L
te <- c(gt[1] +  col_off            * gt[2],   # xmin
        gt[4] + (row_off + height)  * gt[6],   # ymin (gt[6] is negative)
        gt[1] + (col_off + width)   * gt[2],   # xmax
        gt[4] +  row_off            * gt[6])   # ymax

a <- ck_read(paste0(scene, "B04.tif"), te = te)   # native window, no reprojection
dim(a)
str(attributes(a))
```

The array maps onto async-tiff's `RasterArray`: the values are `data`, `dim(a)`
is `shape`, `attr(a, "geotransform")` is `transform`, `attr(a, "crs")` is `crs`,
and `attr(a, "nodata")` stands in for `mask` (derive one with
`a == attr(a, "nodata")`).

## Building a stack from per-band COGs

Sentinel-2 L2A stores **one COG per band**, at each band's native resolution
(10 m for the visible/NIR bands, 20 m for the SWIR bands). To assemble an
analysis-ready stack you read each band onto the *same* grid. Because `ck_read()`
aligns output to the `tr` grid by default, bands fetched with the same `te`/`tr`
come back co-registered --- even when their native resolutions differ, GDAL
resamples each onto the common grid:

```{r stack}
te20 <- c(gt[1] + 4096 * gt[2], gt[4] - (4096 + 512) * abs(gt[6]),
          gt[1] + (4096 + 512) * gt[2], gt[4] - 4096 * abs(gt[6]))

bands <- c(B08 = "B08.tif", B04 = "B04.tif", B11 = "B11.tif")  # 10 m, 10 m, 20 m
layers <- lapply(bands, function(b) {
  ck_read(paste0(scene, b), te = te20, tr = c(20, 20), r = "bilinear")
})
vapply(layers, dim, integer(2))         # identical dims -> co-registered

stack <- array(0, dim = c(dim(layers[[1]]), length(layers)))
for (k in seq_along(layers)) stack[, , k] <- layers[[k]]
dim(stack)
```

(cptkirk's multi-source form, `ck_warp(src = c(tile_a, tile_b), …)`, is a
*spatial* mosaic of the same data across adjacent tiles --- not a band stack, so
combining separate band files is a "read each, bind" as above.)

## Reprojecting on the way out

The same call warps as it reads. GEBCO is global and in EPSG:4326; here we pull
a region of it reprojected to Web Mercator and resampled to 2 km, fetching only
the overview and window that target needs:

```{r gebco, warning = FALSE}
gebco <- "https://projects.pawsey.org.au/idea-gebco-tif/GEBCO_2024.tif"

out <- ck_warp(
  gebco, tempfile(fileext = ".tif"),
  t_srs = "EPSG:3857",
  te    = c(110, -45, 155, -10),   # Australia, in EPSG:4326 ...
  te_srs = "EPSG:4326",            # ... so name the extent's CRS
  tr    = c(2000, 2000),
  r     = "average"
)
ds <- new(gdalraster::GDALRaster, out)
gdalraster::plot_raster(ds, legend = TRUE)
ds$close()
```

The remote read fetched only the tiles covering that extent, at the overview
matching 2 km, then handed the reprojection and resampling to GDAL.
